---
title: "Practice"
author: "Jingyu Gao"
date: "2/18/2018"
output: html_document
---

### <span style="color:#009FDA">Introduction</span>

<span style="color:#002244">This document analyses the global trend of **access to sanitation facilities** data provided by the World Development Indicators database using data from 1990 to 2015. It answers three primary questions:</span>

* How is access to sanitation facilities changing over time?
* Is there an overall trend?
* Does it vary by [income group](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups)?

<span style="color:#002244">Access to sanitation facilities is [defined](https://data.worldbank.org/indicator/SH.STA.ACSN) as **the percentage of the population using improved sanitation facilities**. Improved sanitation facilities are likely to ensure hygienic separation of human excreta from human contact. They include flush/pour flush (to piped sewer system, septic tank, pit latrine), ventilated improved pit (VIP) latrine, pit latrine with slab, and composting toilet.</span>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r message=FALSE, results='hide',warning=FALSE, echo=FALSE}

##############################
# 1. Extract Data and Clean  #
##############################

# Following code is intended to retrieve the improved sanitation facilities (% of population with access) from the World Development Indicators database. It extracts data via WBG API data formatted in JSON, then unpacks the nested JSON web page data into the desired format for analysis.

# Load necessary packages
  library(RJSONIO)
# library(WDI)
  library(plotly)
  library(ggvis)
  library(ggplot2)
  library(jsonlite)
  library(gridExtra)

# Create a dataframe that stores 2-digit iso code, 3-digit iso code, income level for each country
  # This dataframe is intended to merge data with region, income level as separate columns

  # Access the WBG country list via API  
  iso_list <- flatten(data.frame(fromJSON('http://api.worldbank.org/v2/countries?format=json&per_page=500')))
  
  # Keep id (3-digit iso), iso2Code (2-digit iso), region name, income level columns, and individual countries only
  iso_list <- subset(iso_list, region.value!="Aggregates",select=c(id,iso2Code,region.value,region.id, incomeLevel.id, incomeLevel.value))
  
  # Rename the country list columns for merging
  colnames(iso_list) <- c("iso_3","iso_2","region","region_id","income","income_id")

# Create region and income group lists for data query
  # WBG region:
  wbg_region_all <- c("1W","Z7","ZQ","8S","ZJ","ZG","Z4","XU")
  wbg_region_dvpg <- c("1W","8S","7E","XQ","ZF","XJ","4E")
  
  # WBG income level:
  wbg_income <- c("1W","XM","XN","XT","XD")
  
# Find data information: total pages, number of data per page
  page_total <- data.frame(fromJSON('http://api.worldbank.org/v2/countries/all/indicators/SH.STA.ACSN?format=json&date=1990:2015&per_page=500'))[[2]][1]

# Create list of colnames that are relevant to subset 
  keep_col <- c("country.id","country.value","date","value")

# Write a loop to construct complete dataset for sanitation
  san_dat <- data.frame()
  for (i in 1:page_total) {
    # Generate url for each page
    url <- paste0('http://api.worldbank.org/v2/countries/all/indicators/SH.STA.ACSN?format=json&date=1990:2015&per_page=500&page=',i)
    # Get json data from each page and transform it into dataframe
    dat <- data.frame(fromJSON(url)[2], row.names = NULL)
    # Subset and keep relevant columns using keep_col created before
    dat_cl <- flatten(dat)[,keep_col]
    # Append each page and combine them into a single dataset
    san_dat <- rbind(san_dat, dat_cl) }

# Rename the appended dataset
  colnames(san_dat) <- c("iso_2","name","year","sani_value")

# Check the data type of each column and transform
  sapply(san_dat, class); san_dat$year <- as.numeric(san_dat$year)

# Check and delete the missing data
# The dataset actually started only after 1990. According to WBG's note, updates for this data halted
# since 2015. 
  unique(san_dat$year[!is.na(san_dat$sani_value)])
  san_dat <- san_dat[san_dat$year>=1990 & san_dat$year<=2015,]
  
# Merge with iso_list by using 2-digit iso code
  # There are 217 individual countries in dataset
  san_dat_merge <- merge(iso_list, san_dat, by="iso_2")
  length(unique(san_dat_merge$iso_3))

# Delete extra spaces in san_dat_merge data
  # "Latin America & Caribbean " and "Sub-Sahara Africa " have an extra space at the right end 
  unique(san_dat_merge$region) 
  san_dat_merge$region <- trimws(san_dat_merge$region, "both")
  
# Add a column that contains regional regional aggregates for each country observation
  # This is intended to compare the each country with the regional aggregates
  # Note that the reginoal sanitation rate may not be the simple average of sanitation rate of all
  # countries in a region. For accuracy purposes, we'd better use the calculated regional aggregates
  # already existing in WBG API than calculate the aggregates on our own, given the actual methodology
  # of weighted averaging is unknown.
  region_merge <- subset(san_dat, select=c("name","year","sani_value"))
  colnames(region_merge) <- c("region", "year", "region_agg")
  san_dat_merge <- merge(san_dat_merge, region_merge, by = c("region","year"))

  
##############################
# 2. Set General Graph Style #
##############################
  
  # Style
  style <- theme(
    panel.background = element_rect(fill="white"),
    panel.grid.major.y = element_line(colour="grey", size=0.1,linetype = "dashed"),
    axis.line = element_line(colour="grey", size=0.1),
    legend.background=element_rect(fill=NULL,colour = NULL),
    legend.key=element_blank(),
    plot.title = element_text(hjust=0.5),
    axis.title=element_text(face="italic"),
    plot.subtitle = element_text(face="italic",hjust=0.5),
    plot.caption = element_text(hjust=0)
  )
  
  
  
```

### <span style="color:#009FDA">Global Trend 1990-2015</span>

``` {r echo=FALSE}
  # Retrieve calculations to be used in written paragraph

  wld_dvlp <- san_dat[san_dat$iso_2 %in% wbg_region_dvpg,]
  eap_2015 <- round(wld_dvlp[wld_dvlp$iso_2=="4E" &wld_dvlp$year==2015,"sani_value"],2)
  eap_1990 <- round(wld_dvlp[wld_dvlp$iso_2=="4E" &wld_dvlp$year==1990,"sani_value"],2)
  ssa_1990 <- round(wld_dvlp[wld_dvlp$iso_2=="ZF" &wld_dvlp$year==1990,"sani_value"],2)
  ssa_2015 <- round(wld_dvlp[wld_dvlp$iso_2=="ZF" &wld_dvlp$year==2015,"sani_value"],2)
  
  wld_dvlp$region[wld_dvlp$name=="East Asia & Pacific (excluding high income)"] <- "EAP"
  wld_dvlp$region[wld_dvlp$name=="Europe & Central Asia (excluding high income)"] <- "ECA"
  wld_dvlp$region[wld_dvlp$name=="Latin America & Caribbean (excluding high income)"] <- "LAC"
  wld_dvlp$region[wld_dvlp$name=="Middle East & North Africa (excluding high income)"] <- "MENA"
  wld_dvlp$region[wld_dvlp$name=="South Asia"] <- "SA"
  wld_dvlp$region[wld_dvlp$name=="Sub-Saharan Africa (excluding high income)"] <- "SSA"
  wld_dvlp$region[wld_dvlp$name=="World"] <- "World"
```

<span style="color:#002244">Over the period 1990-2015, access to sanitation facilities has been improved steadily, but the level of progress is heterogeneous around the world. East Asia and Pacific (EAP) witnessed the biggest success: over `r eap_2015` percent of its population are now using improved sanitation facilities, substantially increasing from `r eap_1990` percent in 1990. Since 2001, EAP has greater access to sanitation facilities than the world average. At the same time, developing countries in South Asia (SA) and Sub-Sahara Africa (SSA) still are lagging behind the rest of the world, despite significant improvement in SA. Condition in SSA has barely changed: `r ssa_2015` percent of people there have access to improved sanitation facilities in 2015, only slightly higher than what it was in 1990 (`r ssa_1990` percent).</span>

```{r message=FALSE, echo=FALSE, warning =FALSE, fig.align="center"}

# Figure 1: Global Sanitation by Region Over Time
 

  wld_reg_dvlp <- ggplot(wld_dvlp, aes(x=year,y=sani_value,color=region,group=region, text=paste("Region:",region,"<br>","Sanitation:",paste(round(sani_value,1),"%", sep =""),"<br>","Year:",year))) + geom_point(size=1.5)+geom_line(size=1, alpha=0.8) +style + scale_color_manual(name=NULL,values=c("#002244","#00A996","#006068","grey","#98252B","#EB1C2D","#009FDA")) + labs(title="Figure 1. Access to Improved Sanitation Facilities",x="Year",y=NULL, caption="Note: High income countries in these regions and North America are excluded.") + theme(legend.position="right") + annotate("text", x=2000, y=40, label="South Asia and Sub-Sahara Africa \nhave long lagged behind.", hjust=0)

  ggplotly(wld_reg_dvlp,tooltip=c("text"),width=800,height=600, r=0) %>% layout(margin = list(l=150, b=100, t=120), yaxis=list(title= 
      "% of population using\nimproved sanitation facilities"
      ), legend=(list(y=0.5)))

```


<span style="color:#002244">A closer glance at individual countries reveals a more diverse picture. During the period 1995-2015, the top 10 countries with the greatest progress have experienced more than 30 percent points increase in the share of population using improved sanitation facilities. Lao PDR tops all other countries: its coverage expanded from 20.4 to 70.9 percent in the two decades before 2015. However, there are also countries where the situation is worsening. In Djibouti, the indicator shrank to 47.4 from 65.1 percent during the same period, one of the worst deterioration among all countries.</span>


```{r message=FALSE, echo=FALSE, warning =FALSE, fig.align="center"}

# Figure 2 Top 10 countries with greatest increase/decrease in sanitation facilities

  # Construct dataset that calculates difference between 2015 and 1995
  wld_point <- na.omit(subset(san_dat_merge, income != "HIC", select=c("region","income_id","iso_3","name","year","sani_value")))
  wld_point <- wld_point[wld_point$year>=1995 & wld_point$year<=2015,]
  wld_point$year <- as.character(wld_point$year)
  
  # Generate a column that contains san_value for 2015 to sort top countries in the graph
  temp <- wld_point[wld_point$year=="2015",c("iso_3","sani_value")]
  colnames(temp) <- c("iso_3","rank_2015")
  wld_point <- merge(wld_point, temp, by="iso_3")
  wld_point$iso_3 <- factor(wld_point$iso_3, levels=wld_point$iso_3[order(wld_point$rank_2015)])
  
  # Create a dataset that ranks countries by the difference between 1995 and 2015
  rank_dif <- na.omit(reshape(wld_point,
         idvar = c("rank_2015","name","income_id","region","iso_3"),
         timevar =  "year",
         direction = "wide"))
  rank_dif$dif <- rank_dif$sani_value.2015-rank_dif$sani_value.1995
  rank_dif <- rank_dif[,c("iso_3","dif")]
  
  # Merge wld_point with rank_dif, adding the column that contains the change between 2015 and 1995
  wld_point <-merge(wld_point, rank_dif, by="iso_3")
  
  # Group countries by dif into three groups
  wld_point$change[wld_point$dif<=0] <- "4"
  wld_point$change[wld_point$dif>0&wld_point$dif<=15] <- "3"
  wld_point$change[wld_point$dif>15&wld_point$dif<=30] <- "2"
  wld_point$change[wld_point$dif>30] <- "1"
  wld_point$change <- factor(wld_point$change, labels=c("+ (30% point more)","+ (<= 30% point)","+ (<= 15% point)","-- (<= 0% point)"))

  
  # Set customized style
  style2 <- theme(panel.grid.major.x = element_line(colour="grey", size=0.1,linetype = "dashed"))
  
  # Visualize point change over time
  test <- ggplot(wld_point, aes(x=sani_value,y=iso_3,color=change,group=change, frame=year,ids=iso_3, text=paste("Country:",name,"<br>","Sanitation:",paste(sani_value,"%", sep =""),"<br>", "Year:", year,"<br>","95-15 Change:",paste0(round(dif,1),"%")))) + geom_point(size=4.5, alpha=0.8)+ theme_minimal() + scale_colour_manual(values=c("#002244","#04719C","#009FDA","#F05023"),name=NULL) + labs(title="Figure 2. Access to Improved Sanitation Facilities 1995-2015\n(level of 95-15 % point change denoted by color)",y="Country (3-digit ISO code)",x="% of population using \nimproved sanitation facilities")
  
 # text=paste("Country:",name,"<br>","Sanitation:",paste(sani_value,"%", sep =""),"<br>", "Year:", year,"<br>","95-15 Change:",paste0(round(dif,1),"%")),
  
  ggplotly(test, height=2200,tooltip = "text") %>% animation_opts(frame=900, transition=100,redraw=T)  %>% layout(margin = list(l=150, b=100, t=120))
```




### <span style="color:#009FDA">Better Income, Better Sanitation</span>
<span style="color:#002244">As indicated in Figure 3, better income is associated with greater access to facilities that ensure hygienic seperation of human excreta. Almost everyone (over 90 percent of population) living in high income countries have access to such improved sanitation facilities, while in low income countries fewer than half of their population do so. A comparison between 1990 and 2015 also illustrates the fact that when countries become richer, more of their citizens would use improved sanitation facilities, as shown by the colored spots that shifted up- and right-ward in 2015 compared to 1990.</span>

```{r message=FALSE, echo=FALSE, warning =FALSE, fig.align="center"}

# Figure 3: Global Sanitation by Income Group 1990 and 2015 (Scatterplot)

  # Retrieve GNI per capita (NY.GNP.PCAP.CD) used by WBG to define income group
  gnipc <- data.frame()  

  for (j in c(1990, 2015)){
    url <- paste0('http://api.worldbank.org/v2/countries/all/indicators/NY.GNP.PCAP.CD?format=json&date=',j,'&per_page=500')
    assign(paste("gnipc",j,sep="_"),flatten(data.frame(fromJSON(url)))[,c("country.id","date","value")])
                          }

  # Construct data subset to scatterplot gni per capita against access to sanitation facilities 
  gnipc <- rbind(gnipc_1990,gnipc_2015)
  colnames(gnipc) <- c("iso_2","year","gnipc") 
  
  # Merge with sanitation data
  wld_income <- san_dat_merge[san_dat_merge$year==1990 | san_dat_merge$year==2015, c("iso_2","year","income_id","sani_value","name")]
  wld_income <- merge(wld_income, gnipc, by=c("iso_2","year"))
  
  # Generate the log of gni per capita
  wld_income$lggnipc <- log(wld_income$gnipc)
  income <- c("High income", "Upper middle income","Lower middle income","Low income")

  # Scatterplot the graph
  wld_income_sani <- ggplot(wld_income, aes(x=lggnipc,y=sani_value, group = year, color=income_id, text=paste("Country:",name,"<br>","Sanitation:",paste(sani_value,"%", sep =""),"<br>","GNI per capita:",paste0("$",gnipc)))) + geom_point(size=3, alpha=0.8) + facet_wrap(~ year, ncol=2) + style + labs(title="Figure 3. Income Level and Sanitation Facilities",subtitle="Comparison between 1990 and 2015",x="Log of GNI per capita",y=" ") + theme(legend.position="right") + scale_color_manual(name=NULL,limits=income,labels=income,values=c("#002244","#006068","#00A996","grey")) 

  
  ggplotly(wld_income_sani,tooltip=c("text"),width=800,height=600, r=0) %>% layout(margin = list(l=150, b=100, t=120), yaxis=list(title= 
      "% of population using\nimproved sanitation facilities"
      ), legend=(list(y=0.5)))
```


